\section{Device parts}\label{sec:Basic Facilities of a Virtio Device / Device parts}

Device parts represent the device state, with parts for basic
device facilities such as driver features, as well as transport specific
and device type specific parts. In memory, each device part consists
of a header \field{struct virtio_dev_part_hdr} followed by
the device part data in \field{value}. The driver can get and set
these device parts using administration commands.

\begin{lstlisting}
struct virtio_dev_part_hdr {
        le16 part_type;
        u8 flags;
        u8 reserved;
        union {
                struct {
                        le32 offset;
                        le32 reserved;
                } pci_common_cfg;
                struct  {
                        le16 index;
                        u8 reserved[6];
                } vq_index;
        } selector;
        le32 length;
};

#define VIRTIO_DEV_PART_F_OPTIONAL 0

struct virtio_dev_part {
        struct virtio_dev_part_hdr hdr;
        u8 value[];
};

\end{lstlisting}

Each device part consists of a fixed size \field{hdr} followed by optional
part data in field \field{value}. The device parts are divided into
two categories and identified by \field{part_type}. The common device parts are
independent of the device type and, are in the range \field{0x0000 - 0x01FF}. Common
device parts are listed in
\ref{table:Basic Facilities of a Virtio Device / Device parts / Common device parts}
The device parts in the range \field{0x0200 - 0x05FF} are specific to a device type
such as a network or console device.
The device part is identified by the \field{part_type} field as listed:

\begin{description}
\item[0x0000 - 0x01FF] - common part - used to describe a part of the device that
                         is independent of the device type
\item[0x0200 - 0x05FF] - device type specific part - used to indicate parts
                         that are device type specific
\item[0x0600 - 0xFFFF] - reserved
\end{description}

Some device parts are optional, the device can function without them.
For example, such parts can help improve performance, with the device working
slower, yet still correctly, even without the parts. In another example,
optional parts can be used for validation, with the device being able to deduce
the part itself, the part being helpful to detect driver or user errors.
Such device parts are marked optional by setting bit 0
(VIRTIO_DEV_PART_F_OPTIONAL) in the \field{flags}.

\field{reserved} is reserved and set to zero.

\field{length} indicates the length of the \field{value} in bytes. The length
of the device part depends on the device part itself and is described separately.
The device part data is in \field{value} and is \field{part_type} specific.

\field{selector} further specifies the part. It is only used for some
\field{part_type} values.

\field{selector.pci_common_cfg.offset} is the offset of the
field in the \nameref{sec:Virtio Transport Options / Virtio Over PCI Bus / PCI Device Layout / Common configuration structure layout}. It is valid only when the \field{part_type} is set to VIRTIO_DEV_PART_PCI_COMMON_CFG,
otherwise it is reserved and set to 0.

\field{selector.vq_index.index} is the index of the virtqueue. It is valid
only when the \field{part_type} is VIRTIO_DEV_PART_VQ_CFG or
VIRTIO_DEV_PART_VQ_CFG.

\subsection{Common device parts}\label{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts}

Common parts are independent of the device type.
\field{part_type} and \field{value} for each part are documented as follows:

\begin{table}
\caption{Common device parts}
\label{table:Basic Facilities of a Virtio Device / Device parts / Common device parts}
\begin{xltabular}{\textwidth}{ |l||l|X| }
\hline
Type & Name & Description \\
\hline \hline
0x100 & VIRTIO_DEV_PART_DEV_FEATURES & Device features, see \ref{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts / VIRTIO_DEV_PART_DEV_FEATURES} \\
\hline
0x101 & VIRTIO_DEV_PART_DRV_FEATURES & Driver features, \ref{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts / VIRTIO_DEV_PART_DRV_FEATURES} \\
\hline
0x102 & VIRTIO_DEV_PART_PCI_COMMON_CFG & PCI common configuration, see \ref{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts / VIRTIO_DEV_PART_PCI_COMMON_CFG} \\
\hline
0x103 & VIRTIO_DEV_PART_DEVICE_STATUS & Device status, see \ref{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts / VIRTIO_DEV_PART_DEVICE_STATUS} \\
\hline
0x104 & VIRTIO_DEV_PART_VQ_CFG & Virtqueue configuration, see \ref{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts / VIRTIO_DEV_PART_VQ_CFG} \\
\hline
0x105 & VIRTIO_DEV_PART_VQ_NOTIFY_CFG & Virtqueue notification configuration, see \ref{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts / VIRTIO_DEV_PART_VQ_NOTIFY_CFG} \\
\hline
0x106 - 0x2FF & - & Common device parts range reserved for future \\
\hline
\hline
\end{xltabular}
\end{table}

\subsubsection{VIRTIO_DEV_PART_DEV_FEATURES}
\label{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts / VIRTIO_DEV_PART_DEV_FEATURES}

For VIRTIO_DEV_PART_DEV_FEATURES, \field{part_type} is set to 0x100.
The VIRTIO_DEV_PART_DEV_FEATURES field indicates features offered by the device.
\field{value} is in the format of \field{struct virtio_dev_part_features}.
\field{feature_bits} is in the format listed in
\ref{sec:Basic Facilities of a Virtio Device / Feature Bits}.
\field{length} is the length of the \field{struct virtio_dev_part_features}.

If the VIRTIO_DEV_PART_DEV_FEATURES device part is present, there is exactly
one instance of it in the get or set commands.

The VIRTIO_DEV_PART_DEV_FEATURES part is optional for which
the VIRTIO_DEV_PART_F_OPTIONAL (bit 0) \field{flags} is set.

\begin{lstlisting}
struct virtio_dev_part_features {
        le64 feature_bits[];
};
\end{lstlisting}

\subsubsection{VIRTIO_DEV_PART_DRV_FEATURES}
\label{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts / VIRTIO_DEV_PART_DRV_FEATURES}

For VIRTIO_DEV_PART_DRV_FEATURES, \field{part_type} is set to 0x101.
The VIRTIO_DEV_PART_DRV_FEATURES field indicates features set by the driver.
\field{value} is in the format of \field{struct virtio_dev_part_features}.
\field{feature_bits} is in the format listed in
\ref{sec:Basic Facilities of a Virtio Device / Feature Bits}.
\field{length} is the length of the \field{struct virtio_dev_part_features}.

If the VIRTIO_DEV_PART_DEV_FEATURES device part present, there is exactly
one instance of it in the get or set commands.

\subsubsection{VIRTIO_DEV_PART_PCI_COMMON_CFG}
\label{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts / VIRTIO_DEV_PART_PCI_COMMON_CFG}

For VIRTIO_DEV_PART_PCI_COMMON_CFG, \field{part_type} is set to 0x102.
VIRTIO_DEV_PART_PCI_COMMON_CFG refers to the common device configuration
fields. \field{offset} refers to the
byte offset of single field in the common configuration layout described in
\field{struct virtio_pci_common_cfg}. \field{value} is in the format depending on
the \field{offset}, for example when \field{cfg_offset = 18}, \field{value}
is in the format of \field{num_queues}. \field{length} is the length of
\field{value} in bytes of a single structure field whose offset is \field{offset}.

One or multiple VIRTIO_DEV_PART_PCI_COMMON_CFG parts may exist in the
get or set commands; each such part corresponds to a unique \field{offset}.

\subsubsection{VIRTIO_DEV_PART_DEVICE_STATUS}
\label{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts / VIRTIO_DEV_PART_DEVICE_STATUS}

For VIRTIO_DEV_PART_DEVICE_STATUS, \field{part_type} is set to 0x103.
The VIRTIO_DEV_PART_DEVICE_STATUS field indicates the device status as listed in
\ref{sec:Basic Facilities of a Virtio Device / Device Status Field}.
\field{value} is in the format \field{device_status} of
\field{struct virtio_pci_common_cfg}.

If the VIRTIO_DEV_PART_DEV_FEATURES device part is present, there is exactly
one instance of it in the get or set commands.

There is exactly one part may exist in the get or set
commands.

\subsubsection{VIRTIO_DEV_PART_VQ_CFG}
\label{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts / VIRTIO_DEV_PART_VQ_CFG}

For VIRTIO_DEV_PART_VQ_CFG, \field{part_type} is set to 0x104.
\field{value} is in the format \field{struct virtio_dev_part_vq_cfg}.
\field{length} is the length of \field{struct virtio_dev_part_vq_cfg}.

\begin{lstlisting}
struct virtio_dev_part_vq_cfg {
        le16 queue_size;
        le16 vector;
        le16 enabled;
        le16 reserved;
        le64 queue_desc;
        le64 queue_driver;
        le64 queue_device;
};
\end{lstlisting}

\field{queue_size}, \field{vector}, \field{queue_desc},
\field{queue_driver} and \field{queue_device} correspond to the
fields of \field{struct virtio_pci_common_cfg} when used for PCI transport.

One or multiple instances of the device part VIRTIO_DEV_PART_VQ_CFG may exist in
the get and set commands. Each such device part corresponds to a unique virtqueue identified
by the \field{vq_index.index}.

\subsubsection{VIRTIO_DEV_PART_VQ_NOTIFY_CFG}
\label{sec:Basic Facilities of a Virtio Device / Device parts / Common device parts / VIRTIO_DEV_PART_VQ_NOTIFY_CFG}

For VIRTIO_DEV_PART_VQ_NOTIFY_CFG, \field{part_type} is set to 0x105.
\field{value} is in the format \field{struct virtio_dev_part_vq_notify_data}.
\field{length} is the length of \field{struct virtio_dev_part_vq_notify_data}.

\begin{lstlisting}
struct virtio_dev_part_vq_notify_data {
        le16 queue_notify_off;
        le16 queue_notif_config_data;
        u8 reserved[4];
};
\end{lstlisting}

\field{queue_notify_off} and \field{queue_notif_config_data} corresponds to the
fields in \field{struct virtio_pci_common_cfg} described in the
\nameref{sec:Virtio Transport Options / Virtio Over PCI Bus / PCI Device Layout / Common configuration structure layout}.

One or multiple instance of the device part VIRTIO_DEV_PART_VQ_NOTIFY_CFG may exist
in the get and set commands, each such device part corresponds to a unique
virtqueue identified by the \field{vq_index.index}.

\field{reserved} is reserved and set to 0.

\subsection{Assumptions}
For the SR-IOV group type, some hypervisors do not allow the driver to access
the PCI configuration space and the MSI-X Table space directly. Such hypervisors
query and save these fields without the need for this device parts.
Therefore, this version of the specification does not have it in the device parts. A future
extension of the device part may further include them as new device part.
