\section{Virtio Over Channel I/O}\label{sec:Virtio Transport Options / Virtio Over Channel I/O}

S/390 based virtual machines support neither PCI nor MMIO, so a
different transport is needed there.

virtio-ccw uses the standard channel I/O based mechanism used for
the majority of devices on S/390. A virtual channel device with a
special control unit type acts as proxy to the virtio device
(similar to the way virtio-pci uses a PCI device) and
configuration and operation of the virtio device is accomplished
(mostly) via channel commands. This means virtio devices are
discoverable via standard operating system algorithms, and adding
virtio support is mainly a question of supporting a new control
unit type.

As the S/390 is a big endian machine, the data structures transmitted
via channel commands are big-endian: this is made clear by use of
the types be16, be32 and be64.

\subsection{Basic Concepts}\label{sec:Virtio Transport Options / Virtio over channel I/O / Basic Concepts}

As a proxy device, virtio-ccw uses a channel-attached I/O control
unit with a special control unit type (0x3832) and a control unit
model corresponding to the attached virtio device's subsystem
device ID, accessed via a virtual I/O subchannel and a virtual
channel path of type 0x32. This proxy device is discoverable via
normal channel subsystem device discovery (usually a STORE
SUBCHANNEL loop) and answers to the basic channel commands:

\begin{itemize}
\item NO-OPERATION (0x03)
\item BASIC SENSE (0x04)
\item TRANSFER IN CHANNEL (0x08)
\item SENSE ID (0xe4)
\end{itemize}

For a virtio-ccw proxy device, SENSE ID will return the following
information:

\begin{tabular}{ |l|l|l| }
\hline
Bytes & Description & Contents \\
\hline \hline
0     & reserved              & 0xff \\
\hline
1-2   & control unit type     & 0x3832 \\
\hline
3     & control unit model    & <virtio device id> \\
\hline
4-5   & device type           & zeroes (unset) \\
\hline
6     & device model          & zeroes (unset) \\
\hline
7-255 & extended SenseId data & zeroes (unset) \\
\hline
\end{tabular}

A virtio-ccw proxy device facilitates:
\begin{itemize}
\item Discovery and attachment of virtio devices (as described above).
\item Initialization of virtqueues and transport-specific facilities (using
      virtio-specific channel commands).
\item Notifications (via hypercall and a combination of I/O interrupts
      and indicator bits).
\end{itemize}

\subsubsection{Channel Commands for Virtio}\label{sec:Virtio Transport Options / Virtio
over channel I/O / Basic Concepts/ Channel Commands for Virtio}

In addition to the basic channel commands, virtio-ccw defines a
set of channel commands related to configuration and operation of
virtio:

\begin{lstlisting}
#define CCW_CMD_SET_VQ 0x13
#define CCW_CMD_VDEV_RESET 0x33
#define CCW_CMD_SET_IND 0x43
#define CCW_CMD_SET_CONF_IND 0x53
#define CCW_CMD_SET_IND_ADAPTER 0x73
#define CCW_CMD_READ_FEAT 0x12
#define CCW_CMD_WRITE_FEAT 0x11
#define CCW_CMD_READ_CONF 0x22
#define CCW_CMD_WRITE_CONF 0x21
#define CCW_CMD_WRITE_STATUS 0x31
#define CCW_CMD_READ_VQ_CONF 0x32
#define CCW_CMD_SET_VIRTIO_REV 0x83
#define CCW_CMD_READ_STATUS 0x72
\end{lstlisting}

\subsubsection{Notifications}\label{sec:Virtio Transport Options / Virtio
over channel I/O / Basic Concepts/ Notifications}

Available buffer notifications are realized as a hypercall. No additional
setup by the driver is needed. The operation of available buffer
notifications is described in section \ref{sec:Virtio Transport Options /
Virtio over channel I/O / Device Operation / Guest->Host Notification}.

Used buffer notifications are realized either as so-called classic or
adapter I/O interrupts depending on a transport level negotiation. The
initialization is described in sections \ref{sec:Virtio Transport Options
/ Virtio over channel I/O / Device Initialization / Setting Up Indicators
/ Setting Up Classic Queue Indicators} and \ref{sec:Virtio Transport
Options / Virtio over channel I/O / Device Initialization / Setting Up
Indicators / Setting Up Two-Stage Queue Indicators} respectively.  The
operation of each flavor is described in sections \ref{sec:Virtio
Transport Options / Virtio over channel I/O / Device Operation /
Host->Guest Notification / Notification via Classic I/O Interrupts} and
\ref{sec:Virtio Transport Options / Virtio over channel I/O / Device
Operation / Host->Guest Notification / Notification via Adapter I/O
Interrupts} respectively.

Configuration change notifications are done using so-called classic I/O
interrupts. The initialization is described in section \ref{sec:Virtio
Transport Options / Virtio over channel I/O / Device Initialization /
Setting Up Indicators / Setting Up Configuration Change Indicators} and
the operation in section \ref{sec:Virtio Transport Options / Virtio over
channel I/O / Device Operation / Host->Guest Notification / Notification
via Classic I/O Interrupts}.

\devicenormative{\subsubsection}{Basic Concepts}{Virtio Transport Options / Virtio over channel I/O / Basic Concepts}

The virtio-ccw device acts like a normal channel device, as specified
in \hyperref[intro:S390 PoP]{[S390 PoP]} and \hyperref[intro:S390 Common I/O]{[S390 Common I/O]}. In particular:

\begin{itemize}
\item A device MUST post a unit check with command reject for any command
  it does not support.

\item If a driver did not suppress length checks for a channel command,
  the device MUST present a subchannel status as detailed in the
  architecture when the actual length did not match the expected length.

\item If a driver did suppress length checks for a channel command, the
  device MUST present a check condition if the transmitted data does
  not contain enough data to process the command. If the driver submitted
  a buffer that was too long, the device SHOULD accept the command.
\end{itemize}

\drivernormative{\subsubsection}{Basic Concepts}{Virtio Transport Options / Virtio over channel I/O / Basic Concepts}

A driver for virtio-ccw devices MUST check for a control unit
type of 0x3832 and MUST ignore the device type and model.

A driver SHOULD attempt to provide the correct length in a channel
command even if it suppresses length checks for that command.

\subsection{Device Initialization}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization}

virtio-ccw uses several channel commands to set up a device.

\subsubsection{Setting the Virtio Revision}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization / Setting the Virtio Revision}

CCW_CMD_SET_VIRTIO_REV is issued by the driver to set the revision of
the virtio-ccw transport it intends to drive the device with. It uses the
following communication structure:

\begin{lstlisting}
struct virtio_rev_info {
        be16 revision;
        be16 length;
        u8 data[];
};
\end{lstlisting}

\field{revision} contains the desired revision id, \field{length} the length of the
data portion and \field{data} revision-dependent additional desired options.

The following values are supported:

\begin{tabular}{ |l|l|l|l| }
\hline
\field{revision} & \field{length} & \field{data}      & remarks \\
\hline \hline
0        & 0      & <empty>   & legacy interface; transitional devices only \\
\hline
1        & 0      & <empty>   & Virtio 1 \\
\hline
2        & 0      & <empty>   & CCW_CMD_READ_STATUS support \\
\hline
3-n      &        &           & reserved for later revisions \\
\hline
\end{tabular}

Note that a change in the virtio standard does not necessarily
correspond to a change in the virtio-ccw revision.

\devicenormative{\paragraph}{Setting the Virtio Revision}{Virtio Transport Options / Virtio over channel I/O / Device Initialization / Setting the Virtio Revision}

A device MUST post a unit check with command reject for any \field{revision}
it does not support. For any invalid combination of \field{revision}, \field{length}
and \field{data}, it MUST post a unit check with command reject as well. A
non-transitional device MUST reject revision id 0.

A device SHOULD answer with command reject to any virtio-ccw specific
channel command that is not contained in the revision selected by the
driver.

A device MUST answer with command reject to any attempt to select a different revision
after a revision has been successfully selected by the driver.

A device MUST treat the revision as unset from the time the associated
subchannel has been enabled until a revision has been successfully set
by the driver. This implies that revisions are not persistent across
disabling and enabling of the associated subchannel.

\drivernormative{\paragraph}{Setting the Virtio Revision}{Virtio Transport Options / Virtio over channel I/O / Device Initialization / Setting the Virtio Revision}

A driver SHOULD start with trying to set the highest revision it
supports and continue with lower revisions if it gets a command reject.

A driver MUST NOT issue any other virtio-ccw specific channel commands
prior to setting the revision.

After a revision has been successfully selected by the driver, it
MUST NOT attempt to select a different revision.

\paragraph{Legacy Interfaces: A Note on Setting the Virtio Revision}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization / Setting the Virtio Revision / Legacy Interfaces: A Note on Setting the Virtio Revision}

A legacy device will not support the CCW_CMD_SET_VIRTIO_REV and answer
with a command reject. A non-transitional driver MUST stop trying to
operate this device in that case. A transitional driver MUST operate
the device as if it had been able to set revision 0.

A legacy driver will not issue the CCW_CMD_SET_VIRTIO_REV prior to
issuing other virtio-ccw specific channel commands. A non-transitional
device therefore MUST answer any such attempts with a command reject.
A transitional device MUST assume in this case that the driver is a
legacy driver and continue as if the driver selected revision 0. This
implies that the device MUST reject any command not valid for revision
0, including a subsequent CCW_CMD_SET_VIRTIO_REV.

\subsubsection{Configuring a Virtqueue}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization / Configuring a Virtqueue}

CCW_CMD_READ_VQ_CONF is issued by the driver to obtain information
about a queue. It uses the following structure for communicating:

\begin{lstlisting}
struct vq_config_block {
        be16 index;
        be16 max_num;
};
\end{lstlisting}

The requested number of buffers for queue \field{index} is returned in
\field{max_num}.

Afterwards, CCW_CMD_SET_VQ is issued by the driver to inform the
device about the location used for its queue. The transmitted
structure is

\begin{lstlisting}
struct vq_info_block {
        be64 desc;
        be32 res0;
        be16 index;
        be16 num;
        be64 driver;
        be64 device;
};
\end{lstlisting}

\field{desc}, \field{driver} and \field{device} contain the guest
addresses for the descriptor area,
available area and used area for queue \field{index}, respectively. The actual
virtqueue size (number of allocated buffers) is transmitted in \field{num}.

\devicenormative{\paragraph}{Configuring a Virtqueue}{Virtio Transport Options / Virtio over channel I/O / Device Initialization / Configuring a Virtqueue}

\field{res0} is reserved and MUST be ignored by the device.

\paragraph{Legacy Interface: A Note on Configuring a Virtqueue}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization / Configuring a Virtqueue / Legacy Interface: A Note on Configuring a Virtqueue}

For a legacy driver or for a driver that selected revision 0,
CCW_CMD_SET_VQ uses the following communication block:

\begin{lstlisting}
struct vq_info_block_legacy {
        be64 queue;
        be32 align;
        be16 index;
        be16 num;
};
\end{lstlisting}

\field{queue} contains the guest address for queue \field{index}, \field{num} the number of buffers
and \field{align} the alignment. The queue layout follows \ref{sec:Basic Facilities of a Virtio Device / Virtqueues / Legacy Interfaces: A Note on Virtqueue Layout}~\nameref{sec:Basic Facilities of a Virtio Device / Virtqueues / Legacy Interfaces: A Note on Virtqueue Layout}.

\subsubsection{Communicating Status Information}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization / Communicating Status Information}

The driver changes the status of a device via the
CCW_CMD_WRITE_STATUS command, which transmits an 8 bit status
value.

As described in
\ref{devicenormative:Basic Facilities of a Virtio Device / Feature Bits},
a device sometimes fails to set the \field{device status} field: For example, it
might fail to accept the FEATURES_OK status bit during device initialization.

With revision 2, CCW_CMD_READ_STATUS is defined: It reads an 8 bit status
value from the device and acts as a reverse operation to CCW_CMD_WRITE_STATUS.

\drivernormative{\paragraph}{Communicating Status Information}{Virtio Transport Options / Virtio over channel I/O / Device Initialization / Communicating Status Information}

If the device posts a unit check with command reject in response to the
CCW_CMD_WRITE_STATUS command, the driver MUST assume that the device failed
to set the status and the \field{device status} field retained
its previous value.

If at least revision 2 has been negotiated, the driver SHOULD use the
CCW_CMD_READ_STATUS command to retrieve the \field{device status} field after
a configuration change has been detected.

If not at least revision 2 has been negotiated, the driver MUST NOT attempt
to issue the CCW_CMD_READ_STATUS command.

\devicenormative{\paragraph}{Communicating Status Information}{Virtio Transport Options / Virtio over channel I/O / Device Initialization / Communicating Status Information}

If the device fails to set the \field{device status} field
to the value written by the driver, the device MUST assure
that the \field{device status} field is left unchanged and
MUST post a unit check with command reject.

If at least revision 2 has been negotiated, the device MUST return the
current \field{device status} field if the CCW_CMD_READ_STATUS
command is issued.

\subsubsection{Handling Device Features}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization / Handling Device Features}

Feature bits are arranged in an array of 32 bit values, making
for a total of 8192 feature bits. Feature bits are in
little-endian byte order.

The CCW commands dealing with features use the following
communication block:

\begin{lstlisting}
struct virtio_feature_desc {
        le32 features;
        u8 index;
};
\end{lstlisting}

\field{features} are the 32 bits of features currently accessed, while
\field{index} describes which of the feature bit values is to be
accessed. No padding is added at the end of the structure, it is
exactly 5 bytes in length.

The guest obtains the device's device feature set via the
CCW_CMD_READ_FEAT command. The device stores the features at \field{index}
to \field{features}.

For communicating its supported features to the device, the driver
uses the CCW_CMD_WRITE_FEAT command, denoting a \field{features}/\field{index}
combination.

\subsubsection{Device Configuration}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization / Device Configuration}

The device's configuration space is located in host memory.

To obtain information from the configuration space, the driver
uses CCW_CMD_READ_CONF, specifying the guest memory for the device
to write to.

For changing configuration information, the driver uses
CCW_CMD_WRITE_CONF, specifying the guest memory for the device to
read from.

In both cases, the complete configuration space is transmitted.  This
allows the driver to compare the new configuration space with the old
version, and keep a generation count internally whenever it changes.

\subsubsection{Setting Up Indicators}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization / Setting Up Indicators}

In order to set up the indicator bits for host->guest notification,
the driver uses different channel commands depending on whether it
wishes to use traditional I/O interrupts tied to a subchannel or
adapter I/O interrupts for virtqueue notifications. For any given
device, the two mechanisms are mutually exclusive.

For the configuration change indicators, only a mechanism using
traditional I/O interrupts is provided, regardless of whether
traditional or adapter I/O interrupts are used for virtqueue
notifications.

\paragraph{Setting Up Classic Queue Indicators}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization / Setting Up Indicators / Setting Up Classic Queue Indicators}

Indicators for notification via classic I/O interrupts are contained
in a 64 bit value per virtio-ccw proxy device.

To communicate the location of the indicator bits for host->guest
notification, the driver uses the CCW_CMD_SET_IND command,
pointing to a location containing the guest address of the
indicators in a 64 bit value.

If the driver has already set up two-staged queue indicators via the
CCW_CMD_SET_IND_ADAPTER command, the device MUST post a unit check
with command reject to any subsequent CCW_CMD_SET_IND command.

\paragraph{Setting Up Configuration Change Indicators}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization / Setting Up Indicators / Setting Up Configuration Change Indicators}

Indicators for configuration change host->guest notification are
contained in a 64 bit value per virtio-ccw proxy device.

To communicate the location of the indicator bits used in the
configuration change host->guest notification, the driver issues the
CCW_CMD_SET_CONF_IND command, pointing to a location containing the
guest address of the indicators in a 64 bit value.

\paragraph{Setting Up Two-Stage Queue Indicators}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization / Setting Up Indicators / Setting Up Two-Stage Queue Indicators}

Indicators for notification via adapter I/O interrupts consist of
two stages:
\begin{itemize}
\item a summary indicator byte covering the virtqueues for one or more
  virtio-ccw proxy devices
\item a set of contiguous indicator bits for the virtqueues for a
  virtio-ccw proxy device
\end{itemize}

To communicate the location of the summary and queue indicator bits,
the driver uses the CCW_CMD_SET_IND_ADAPTER command with the following
payload:

\begin{lstlisting}
struct virtio_thinint_area {
        be64 summary_indicator;
        be64 indicator;
        be64 bit_nr;
        u8 isc;
} __attribute__ ((packed));
\end{lstlisting}

\field{summary_indicator} contains the guest address of the 8 bit summary
indicator.
\field{indicator} contains the guest address of an area wherein the indicators
for the devices are contained, starting at \field{bit_nr}, one bit per
virtqueue of the device. Bit numbers start at the left, i.e. the most
significant bit in the first byte is assigned the bit number 0.
\field{isc} contains the I/O interruption subclass to be used for the adapter
I/O interrupt. It MAY be different from the isc used by the proxy
virtio-ccw device's subchannel.
No padding is added at the end of the structure, it is exactly 25 bytes
in length.


\devicenormative{\subparagraph}{Setting Up Two-Stage Queue Indicators}{Virtio Transport Options / Virtio over channel I/O / Device Initialization / Setting Up Indicators / Setting Up Two-Stage Queue Indicators}
If the driver has already set up classic queue indicators via the
CCW_CMD_SET_IND command, the device MUST post a unit check with
command reject to any subsequent CCW_CMD_SET_IND_ADAPTER command.

\paragraph{Legacy Interfaces: A Note on Setting Up Indicators}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Initialization / Setting Up Indicators / Legacy Interfaces: A Note on Setting Up Indicators}

In some cases, legacy devices will only support classic queue indicators;
in that case, they will reject CCW_CMD_SET_IND_ADAPTER as they don't know that
command. Some legacy devices will support two-stage queue indicators, though,
and a driver will be able to successfully use CCW_CMD_SET_IND_ADAPTER to set
them up.

\subsection{Device Operation}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Operation}

\subsubsection{Host->Guest Notification}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Operation / Host->Guest Notification}

There are two modes of operation regarding host->guest notification,
classic I/O interrupts and adapter I/O interrupts. The mode to be
used is determined by the driver by using CCW_CMD_SET_IND respectively
CCW_CMD_SET_IND_ADAPTER to set up queue indicators.

For configuration changes, the driver always uses classic I/O
interrupts.

\paragraph{Notification via Classic I/O Interrupts}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Operation / Host->Guest Notification / Notification via Classic I/O Interrupts}

If the driver used the CCW_CMD_SET_IND command to set up queue
indicators, the device will use classic I/O interrupts for
host->guest notification about virtqueue activity.

For notifying the driver of virtqueue buffers, the device sets the
corresponding bit in the guest-provided indicators. If an
interrupt is not already pending for the subchannel, the device
generates an unsolicited I/O interrupt.

If the device wants to notify the driver about configuration
changes, it sets bit 0 in the configuration indicators and
generates an unsolicited I/O interrupt, if needed. This also
applies if adapter I/O interrupts are used for queue notifications.

\paragraph{Notification via Adapter I/O Interrupts}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Operation / Host->Guest Notification / Notification via Adapter I/O Interrupts}

If the driver used the CCW_CMD_SET_IND_ADAPTER command to set up
queue indicators, the device will use adapter I/O interrupts for
host->guest notification about virtqueue activity.

For notifying the driver of virtqueue buffers, the device sets the
bit in the guest-provided indicator area at the corresponding offset.
The guest-provided summary indicator is set to 0x01. An adapter I/O
interrupt for the corresponding interruption subclass is generated.

The recommended way to process an adapter I/O interrupt by the driver
is as follows:

\begin{itemize}
\item Process all queue indicator bits associated with the summary indicator.
\item Clear the summary indicator, performing a synchronization (memory
barrier) afterwards.
\item Process all queue indicator bits associated with the summary indicator
again.
\end{itemize}

\devicenormative{\subparagraph}{Notification via Adapter I/O Interrupts}{Virtio Transport Options / Virtio over channel I/O / Device Operation / Host->Guest Notification / Notification via Adapter I/O Interrupts}

The device SHOULD only generate an adapter I/O interrupt if the
summary indicator had not been set prior to notification.

\drivernormative{\subparagraph}{Notification via Adapter I/O Interrupts}{Virtio Transport Options / Virtio over channel I/O / Device Operation / Host->Guest Notification / Notification via Adapter I/O Interrupts}
The driver
MUST clear the summary indicator after receiving an adapter I/O
interrupt before it processes the queue indicators.

\paragraph{Legacy Interfaces: A Note on Host->Guest Notification}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Operation / Host->Guest Notification / Legacy Interfaces: A Note on Host->Guest Notification}

As legacy devices and drivers support only classic queue indicators,
host->guest notification will always be done via classic I/O interrupts.

\subsubsection{Guest->Host Notification}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Operation / Guest->Host Notification}

For notifying the device of virtqueue buffers, the driver
unfortunately can't use a channel command (the asynchronous
characteristics of channel I/O interact badly with the host block
I/O backend). Instead, it uses a diagnose 0x500 call with subcode
3 specifying the queue, as follows:

\begin{tabular}{ |l|l|l| }
\hline
GPR  &   Input Value     & Output Value \\
\hline \hline
  1   &       0x3         &              \\
\hline
  2   &  Subchannel ID    & Host Cookie  \\
\hline
  3   & Notification data &              \\
\hline
  4   &   Host Cookie     &              \\
\hline
\end{tabular}

When VIRTIO_F_NOTIFICATION_DATA has not been negotiated,
the \field{Notification data} contains the Virtqueue number.

When VIRTIO_F_NOTIFICATION_DATA has been negotiated,
the value has the following format:
\lstinputlisting{notifications-be.c}

See \ref{sec:Basic Facilities of a Virtio Device / Driver notifications}~\nameref{sec:Basic Facilities of a Virtio Device / Driver notifications}
for the definition of the components.

\devicenormative{\paragraph}{Guest->Host Notification}{Virtio Transport Options / Virtio over channel I/O / Device Operation / Guest->Host Notification}
The device MUST ignore bits 0-31 (counting from the left) of GPR2.
This aligns passing the subchannel ID with the way it is passed
for the existing I/O instructions.

The device MAY return a 64-bit host cookie in GPR2 to speed up the
notification execution.

\drivernormative{\paragraph}{Guest->Host Notification}{Virtio Transport Options / Virtio over channel I/O / Device Operation / Guest->Host Notification}

For each notification, the driver SHOULD use GPR4 to pass the host cookie received in GPR2 from the previous notification.

\begin{note}
For example:
\begin{lstlisting}
info->cookie = do_notify(schid,
                         virtqueue_get_queue_index(vq),
                         info->cookie);
\end{lstlisting}
\end{note}

\subsubsection{Resetting Devices}\label{sec:Virtio Transport Options / Virtio over channel I/O / Device Operation / Resetting Devices}

In order to reset a device, a driver sends the
CCW_CMD_VDEV_RESET command. This command does not carry any payload.

The device signals completion of the virtio reset operation through successful
conclusion of the CCW_CMD_VDEV_RESET channel command. In particular, the
command not only triggers the reset operation, but the reset operation is
already completed when the operation concludes successfully.

\devicenormative{\paragraph}{Resetting Devices}{Virtio Transport Options / Virtio over channel I/O / Device Operation / Resetting Devices}

The device MUST finish the virtio reset operation and reinitialize
\field{device status} to zero before it concludes the CCW_CMD_VDEV_RESET
command successfully.

The device MUST NOT send notifications or interact with the queues after
it signaled successful conclusion of the CCW_CMD_VDEV_RESET command.

\drivernormative{\paragraph}{Resetting Devices}{Virtio Transport Options / Virtio over channel I/O / Device Operation / Resetting Devices}

The driver MAY consider the virtio reset operation to be complete already after
successful conclusion of the CCW_CMD_VDEV_RESET channel command, although it
MAY also choose to verify reset completion by reading \field{device status} via
CCW_CMD_READ_STATUS and checking whether it is 0 afterwards.

\subsection{Features reserved for future use}\label{sec:Virtio Transport Options / Virtio over channel I/O / Features reserved for future use}

Devices and drivers utilizing Virtio over channel I/O
do not support the following features:
\begin{itemize}

\item VIRTIO_F_ADMIN_VQ
\item VIRTIO_F_RING_RESET
\item Shared memory regions including VIRTIO_PMEM_F_SHMEM_REGION

\end{itemize}

These features are reserved for future use.
