\section{Traditional Memory Balloon Device}\label{sec:Device Types / Memory Balloon Device}

This is the traditional balloon device.  The device number 13 is
reserved for a new memory balloon interface, with different
semantics, which is expected in a future version of the standard.

The traditional virtio memory balloon device is a primitive device for
managing guest memory: the device asks for a certain amount of
memory, and the driver supplies it (or withdraws it, if the device
has more than it asks for). This allows the guest to adapt to
changes in allowance of underlying physical memory. If the
feature is negotiated, the device can also be used to communicate
guest memory statistics to the host.

\subsection{Device ID}\label{sec:Device Types / Memory Balloon Device / Device ID}
  5

\subsection{Virtqueues}\label{sec:Device Types / Memory Balloon Device / Virtqueues}
\begin{description}
\item[0] inflateq
\item[1] deflateq
\item[2] statsq
\item[3] free_page_vq
\item[4] reporting_vq
\end{description}

  statsq only exists if VIRTIO_BALLOON_F_STATS_VQ is set.

  free_page_vq only exists if VIRTIO_BALLOON_F_FREE_PAGE_HINT is set.

  reporting_vq only exists if VIRTIO_BALLOON_F_PAGE_REPORTING is set.

\subsection{Feature bits}\label{sec:Device Types / Memory Balloon Device / Feature bits}
\begin{description}
\item[VIRTIO_BALLOON_F_MUST_TELL_HOST (0)] Host has to be told before
    pages from the balloon are used.

\item[VIRTIO_BALLOON_F_STATS_VQ (1)] A virtqueue for reporting guest
    memory statistics is present.
\item[VIRTIO_BALLOON_F_DEFLATE_ON_OOM (2) ] Deflate balloon on
    guest out of memory condition.
\item[ VIRTIO_BALLOON_F_FREE_PAGE_HINT(3) ] The device has support for free
    page hinting. A virtqueue for providing hints as to what memory is
    currently free is present. Configuration field \field{free_page_hint_cmd_id}
    is valid.
\item[ VIRTIO_BALLOON_F_PAGE_POISON(4) ] A hint to the device, that the driver
    will immediately write \field{poison_val} to pages after deflating them.
    Configuration field \field{poison_val} is valid.
\item[ VIRTIO_BALLOON_F_PAGE_REPORTING(5) ] The device has support for free
    page reporting. A virtqueue for reporting free guest memory is present.

\end{description}

\drivernormative{\subsubsection}{Feature bits}{Device Types / Memory Balloon Device / Feature bits}
The driver SHOULD accept the VIRTIO_BALLOON_F_MUST_TELL_HOST
feature if offered by the device.

The driver SHOULD clear the VIRTIO_BALLOON_F_PAGE_POISON flag if it will
not immediately write \field{poison_val} to deflated pages (e.g., to
initialize them, or fill them with a poison value).

If the driver is expecting the pages to retain some initialized value,
it MUST NOT accept VIRTIO_BALLOON_F_PAGE_REPORTING unless it also
negotiates VIRTIO_BALLOON_F_PAGE_POISON.

\devicenormative{\subsubsection}{Feature bits}{Device Types / Memory Balloon Device / Feature bits}
If the device offers the VIRTIO_BALLOON_F_MUST_TELL_HOST feature
bit, and if the driver did not accept this feature bit, the
device MAY signal failure by failing to set FEATURES_OK
\field{device status} bit when the driver writes it.
\subparagraph{Legacy Interface: Feature bits}\label{sec:Device
Types / Memory Balloon Device / Feature bits / Legacy Interface:
Feature bits}
As the legacy interface does not have a way to gracefully report feature
negotiation failure, when using the legacy interface,
transitional devices MUST support guests which do not negotiate
VIRTIO_BALLOON_F_MUST_TELL_HOST feature, and SHOULD
allow guest to use memory before notifying host if
VIRTIO_BALLOON_F_MUST_TELL_HOST is not negotiated.

\subsection{Device configuration layout}\label{sec:Device Types / Memory Balloon Device / Device configuration layout}
  \field{num_pages} and \field{actual} are always available.

  \field{free_page_hint_cmd_id} is available if
    VIRTIO_BALLOON_F_FREE_PAGE_HINT has been negotiated. The field is
    read-only by the driver.
  \field{poison_val} is available if VIRTIO_BALLOON_F_PAGE_POISON has been
    negotiated.

\begin{lstlisting}
struct virtio_balloon_config {
        le32 num_pages;
        le32 actual;
        le32 free_page_hint_cmd_id;
        le32 poison_val;
};
\end{lstlisting}

\subparagraph{Legacy Interface: Device configuration layout}\label{sec:Device Types / Memory Balloon Device / Device
configuration layout / Legacy Interface: Device configuration layout}
When using the legacy interface, transitional devices and drivers
MUST format the fields in struct virtio_balloon_config
according to the little-endian format.
\begin{note}
This is unlike the usual convention that legacy device fields are guest endian.
\end{note}

\subsection{Device Initialization}\label{sec:Device Types / Memory Balloon Device / Device Initialization}

The device initialization process is outlined below:

\begin{enumerate}
\item The inflate and deflate virtqueues are identified.

\item If the VIRTIO_BALLOON_F_STATS_VQ feature bit is negotiated:
  \begin{enumerate}
  \item Identify the stats virtqueue.
  \item Add one empty buffer to the stats virtqueue.
  \end{enumerate}

\item If the VIRTIO_BALLOON_F_FREE_PAGE_HINT feature bit is negotiated,
  identify the free_page_vq.
\item If the VIRTIO_BALLOON_F_PAGE_POISON feature bit is negotiated, update
  the \field{poison_val} configuration field.
\item If the VIRTIO_BALLOON_F_PAGE_REPORTING feature bit is negotiated,
  identify the reporting_vq.

\item DRIVER_OK is set: device operation begins.

\item If the VIRTIO_BALLOON_F_STATS_VQ feature bit is negotiated, then
  notify the device about the stats virtqueue buffer.
\item If the VIRTIO_BALLOON_F_PAGE_REPORTING feature bit is negotiated, then
  begin reporting free pages to the device.
\end{enumerate}

\subsection{Device Operation}\label{sec:Device Types / Memory Balloon Device / Device Operation}

The device is driven either by the receipt of a configuration
change notification, or by changing guest memory needs, such as
performing memory compaction or responding to out of memory
conditions.

\begin{enumerate}
\item \field{num_pages} configuration field is examined. If this is
  greater than the \field{actual} number of pages, the balloon wants
  more memory from the guest.  If it is less than \field{actual},
  the balloon doesn't need it all.

\item To supply memory to the balloon (aka. inflate):
  \begin{enumerate}
  \item The driver constructs an array of addresses of unused memory
    pages. These addresses are divided by 4096\footnote{This is historical, and independent of the guest page size.
} and the descriptor
    describing the resulting 32-bit array is added to the inflateq.
  \end{enumerate}

\item To remove memory from the balloon (aka. deflate):
  \begin{enumerate}
  \item The driver constructs an array of addresses of memory pages
    it has previously given to the balloon, as described above.
    This descriptor is added to the deflateq.

  \item If the VIRTIO_BALLOON_F_MUST_TELL_HOST feature is negotiated, the
    guest informs the device of pages before it uses them.

  \item Otherwise, the guest is allowed to re-use pages previously
    given to the balloon before the device has acknowledged their
    withdrawal\footnote{In this case, deflation advice is merely a courtesy.
}.
  \end{enumerate}

\item In either case, the device acknowledges inflate and deflate
requests by using the descriptor.
\item Once the device has acknowledged the inflation or
  deflation, the driver updates \field{actual} to reflect the new number of pages in the balloon.
\end{enumerate}

\drivernormative{\subsubsection}{Device Operation}{Device Types / Memory Balloon Device / Device Operation}
The driver SHOULD supply pages to the balloon when \field{num_pages} is
greater than the actual number of pages in the balloon.

The driver MAY use pages from the balloon when \field{num_pages} is
less than the actual number of pages in the balloon.

The driver MAY supply pages to the balloon when \field{num_pages} is
greater than or equal to the actual number of pages in the balloon.

If VIRTIO_BALLOON_F_DEFLATE_ON_OOM has not been negotiated, the
driver MUST NOT use pages from the balloon when \field{num_pages}
is less than or equal to the actual number of pages in the
balloon.

If VIRTIO_BALLOON_F_DEFLATE_ON_OOM has been negotiated, the
driver MAY use pages from the balloon when \field{num_pages}
is less than or equal to the actual number of pages in the
balloon if this is required for system stability
(e.g. if memory is required by applications running within
 the guest).

The driver MUST use the deflateq to inform the device of pages that it
wants to use from the balloon.

If the VIRTIO_BALLOON_F_MUST_TELL_HOST feature is negotiated, the
driver MUST NOT use pages from the balloon until
the device has acknowledged the deflate request.

Otherwise, if the VIRTIO_BALLOON_F_MUST_TELL_HOST feature is not
negotiated, the driver MAY begin to re-use pages previously
given to the balloon before the device has acknowledged the
deflate request.

In any case, the driver MUST NOT use pages from the balloon
after adding the pages to the balloon, but before the device has
acknowledged the inflate request.

The driver MUST NOT request deflation of pages in
the balloon before the device has acknowledged the inflate
request.

The driver MUST update \field{actual} after changing the number
of pages in the balloon.

The driver MAY update \field{actual} once after multiple
inflate and deflate operations.

\devicenormative{\subsubsection}{Device Operation}{Device Types / Memory Balloon Device / Device Operation}

The device MAY modify the contents of a page in the balloon
after detecting its physical number in an inflate request
and before acknowledging the inflate request by using the inflateq
descriptor.

If the VIRTIO_BALLOON_F_MUST_TELL_HOST feature is negotiated, the
device MAY modify the contents of a page in the balloon
after detecting its physical number in an inflate request
and before detecting its physical number in a deflate request
and acknowledging the deflate request.

\paragraph{Legacy Interface: Device Operation}\label{sec:Device
Types / Memory Balloon Device / Device Operation / Legacy
Interface: Device Operation}
When using the legacy interface, the driver SHOULD ignore the
used length values.
\begin{note}
Historically, some devices put the total descriptor length there,
even though no data was actually written.
\end{note}
When using the legacy interface, the driver MUST write out all
4 bytes each time it updates the \field{actual} value in the
configuration space, using a single atomic operation.

When using the legacy interface, the device SHOULD NOT use the
\field{actual} value written by the driver in the configuration
space, until the last, most-significant byte of the value has been
written.
\begin{note}
Historically, devices used the \field{actual} value, even though
when using Virtio Over PCI Bus the device-specific configuration
space was not guaranteed to be atomic. Using intermediate
values during update by driver is best avoided, except for
debugging.

Historically, drivers using Virtio Over PCI Bus wrote the
\field{actual} value by using multiple single-byte writes in
order, from the least-significant to the most-significant value.
\end{note}
\subsubsection{Memory Statistics}\label{sec:Device Types / Memory Balloon Device / Device Operation / Memory Statistics}

The stats virtqueue is atypical because communication is driven
by the device (not the driver). The channel becomes active at
driver initialization time when the driver adds an empty buffer
and notifies the device. A request for memory statistics proceeds
as follows:

\begin{enumerate}
\item The device uses the buffer and sends a used buffer notification.

\item The driver pops the used buffer and discards it.

\item The driver collects memory statistics and writes them into a
  new buffer.

\item The driver adds the buffer to the virtqueue and notifies the
  device.

\item The device pops the buffer (retaining it to initiate a
  subsequent request) and consumes the statistics.
\end{enumerate}

  Within the buffer, statistics are an array of 10-byte entries.
  Each statistic consists of a 16 bit
  tag and a 64 bit value. All statistics are optional and the
  driver chooses which ones to supply. To guarantee backwards
  compatibility, devices omit unsupported statistics.

\begin{lstlisting}
struct virtio_balloon_stat {
#define VIRTIO_BALLOON_S_SWAP_IN  0
#define VIRTIO_BALLOON_S_SWAP_OUT 1
#define VIRTIO_BALLOON_S_MAJFLT   2
#define VIRTIO_BALLOON_S_MINFLT   3
#define VIRTIO_BALLOON_S_MEMFREE  4
#define VIRTIO_BALLOON_S_MEMTOT   5
#define VIRTIO_BALLOON_S_AVAIL    6
#define VIRTIO_BALLOON_S_CACHES   7
#define VIRTIO_BALLOON_S_HTLB_PGALLOC 8
#define VIRTIO_BALLOON_S_HTLB_PGFAIL  9
        le16 tag;
        le64 val;
} __attribute__((packed));
\end{lstlisting}

\drivernormative{\paragraph}{Memory Statistics}{Device Types / Memory Balloon Device / Device Operation / Memory Statistics}
Normative statements in this section apply if and only if the
VIRTIO_BALLOON_F_STATS_VQ feature has been negotiated.

The driver MUST make at most one buffer available to the device
in the statsq, at all times.

After initializing the device, the driver MUST make an output
buffer available in the statsq.

Upon detecting that device has used a buffer in the statsq, the
driver MUST make an output buffer available in the statsq.

Before making an output buffer available in the statsq, the
driver MUST initialize it, including one struct
virtio_balloon_stat entry for each statistic that it supports.

Driver MUST use an output buffer size which is a multiple of 6
bytes for all buffers submitted to the statsq.

Driver MAY supply struct virtio_balloon_stat entries in the
output buffer submitted to the statsq in any order, without
regard to \field{tag} values.

Driver MAY supply a subset of all statistics in the output buffer
submitted to the statsq.

Driver MUST supply the same subset of statistics in all buffers
submitted to the statsq.

\devicenormative{\paragraph}{Memory Statistics}{Device Types / Memory Balloon Device / Device Operation / Memory Statistics}
Normative statements in this section apply if and only if  the
VIRTIO_BALLOON_F_STATS_VQ feature has been negotiated.

Within an output buffer submitted to the statsq,
the device MUST ignore entries with \field{tag} values that it does not recognize.

Within an output buffer submitted to the statsq,
the device MUST accept struct virtio_balloon_stat entries in any
order without regard to \field{tag} values.

\paragraph{Legacy Interface: Memory Statistics}\label{sec:Device Types / Memory Balloon Device / Device Operation / Memory Statistics / Legacy Interface: Memory Statistics}

When using the legacy interface, transitional devices and drivers
MUST format the fields in struct virtio_balloon_stat
according to the native endian of the guest rather than
(necessarily when not using the legacy interface) little-endian.

When using the legacy interface,
the device SHOULD ignore all values in the first buffer in the
statsq supplied by the driver after device initialization.
\begin{note}
Historically, drivers supplied an uninitialized buffer in the
first buffer.
\end{note}

\subsubsection{Memory Statistics Tags}\label{sec:Device Types / Memory Balloon Device / Device Operation / Memory Statistics Tags}

\begin{description}
\item[VIRTIO_BALLOON_S_SWAP_IN (0)] The amount of memory that has been
  swapped in (in bytes).

\item[VIRTIO_BALLOON_S_SWAP_OUT (1)] The amount of memory that has been
  swapped out to disk (in bytes).

\item[VIRTIO_BALLOON_S_MAJFLT (2)] The number of major page faults that
  have occurred.

\item[VIRTIO_BALLOON_S_MINFLT (3)] The number of minor page faults that
  have occurred.

\item[VIRTIO_BALLOON_S_MEMFREE (4)] The amount of memory not being used
  for any purpose (in bytes).

\item[VIRTIO_BALLOON_S_MEMTOT (5)] The total amount of memory available
  (in bytes).

\item[VIRTIO_BALLOON_S_AVAIL (6)] An estimate of how much memory is available
  (in bytes) for starting new applications, without pushing the system to swap.

\item[VIRTIO_BALLOON_S_CACHES (7)] The amount of memory, in bytes, that can be
  quickly reclaimed without additional I/O. Typically these pages are used for
  caching files from disk.

\item[VIRTIO_BALLOON_S_HTLB_PGALLOC (8)] The number of successful hugetlb page
  allocations in the guest.

\item[VIRTIO_BALLOON_S_HTLB_PGFAIL (9)] The number of failed hugetlb page
  allocations in the guest.
\end{description}

\subsubsection{Free Page Hinting}\label{sec:Device Types / Memory Balloon Device / Device Operation / Free Page Hinting}

Free page hinting is designed to be used during migration to determine what
pages within the guest are currently unused so that they can be skipped over
while migrating the guest. The device will indicate that it is ready to start
performing hinting by setting the \field{free_page_hint_cmd_id} to one of the
non-reserved values that can be used as a command ID. The following values
are reserved:

\begin{description}
\item[VIRTIO_BALLOON_CMD_ID_STOP (0)] Any command ID previously supplied by
  the device is invalid. The driver should stop hinting free pages until a
  new command ID is supplied, but should not release any hinted pages for
  use by the guest.

\item[VIRTIO_BALLOON_CMD_ID_DONE (1)] Any command ID previously supplied by
  the device is invalid. The driver should stop hinting free pages, and
  should release all hinted pages for use by the guest.
\end{description}

When a hint is provided by the driver it indicates that the data contained
in the given page is no longer needed and can be discarded. If the driver
writes to the page this overrides the hint and the data will be retained.
The contents of any stale pages that have not been written to since the
page was hinted may be lost, and if read the contents of such pages will
be uninitialized memory.

A request for free page hinting proceeds as follows:

\begin{enumerate}

\item The driver examines the \field{free_page_hint_cmd_id} configuration field.
  If it contains a non-reserved value then free page hinting will begin.

\item To supply free page hints:
  \begin{enumerate}
  \item The driver constructs an output buffer containing the new value from
    the \field{free_page_hint_cmd_id} configuration field and adds it to the
    free_page_vq.
  \item The driver maps a series of pages and adds them to the
    free_page_vq as individual scatter-gather input buffer entries.
  \item When the driver is no longer able to fetch additional pages to add
    to the free_page_vq, it will construct an output buffer containing the
    command ID VIRTIO_BALLOON_CMD_ID_STOP.
  \end{enumerate}

\item A round of hinting ends either when the driver is no longer able to
  supply more pages for hinting as described above, or when the device
  updates \field{free_page_hint_cmd_id} configuration field to contain either
  VIRTIO_BALLOON_CMD_ID_STOP or VIRTIO_BALLOON_CMD_ID_DONE.

\item The device may follow VIRTIO_BALLOON_CMD_ID_STOP with a new
  non-reserved value for the \field{free_page_hint_cmd_id} configuration
  field in which case it will resume supplying free page hints.

\item Otherwise, if the device provides VIRTIO_BALLOON_CMD_ID_DONE then
  hinting is complete and the driver may release all previously hinted
  pages for use by the guest.

\end{enumerate}

\drivernormative{\paragraph}{Free Page Hinting}{Device Types / Memory Balloon Device / Device Operation / Free Page Hinting}

Normative statements in this section apply if the
VIRTIO_BALLOON_F_FREE_PAGE_HINT feature has been negotiated.

The driver MUST use an output buffer size of 4 bytes for all output
buffers submitted to the free_page_vq.

The driver MUST start hinting by providing an output buffer containing
the current command ID for the given block of pages.

The driver MUST NOT provide more than one output buffer containing the
current command ID.

The driver SHOULD supply pages to the free_page_vq as input buffers when
\field{free_page_hint_cmd_id} specifies a value of 2 or greater.

The driver SHOULD stop supplying pages for hinting when
\field{free_page_hint_cmd_id} specifies a value of VIRTIO_BALLOON_CMD_ID_STOP
or VIRTIO_BALLOON_CMD_ID_DONE.

If the driver is unable to supply pages, it MUST complete hinting by adding
an output buffer containing the command ID VIRTIO_BALLOON_CMD_ID_STOP.

The driver MAY release hinted pages for use by the guest including when the
device has not yet used the descriptor containing the hinting request.

The driver MUST treat the content of all hinted pages as uninitialized
memory.

The driver MUST initialize the contents of any previously hinted page
released before \field{free_page_hint_cmd_id} specifies a value of
VIRTIO_BALLOON_CMD_ID_DONE.

The driver SHOULD release all previously hinted pages once
\field{free_page_hint_cmd_id} specifies a value of
VIRTIO_BALLOON_CMD_ID_DONE.

\devicenormative{\paragraph}{Free Page Hinting}{Device Types / Memory Balloon Device / Device Operation / Free Page Hinting}

Normative statements in this section apply if the
VIRTIO_BALLOON_F_FREE_PAGE_HINT feature has been negotiated.

The device SHOULD set \field{free_page_hint_cmd_id} to
VIRTIO_BALLOON_CMD_ID_STOP any time that it will not be able to make use
of the hints provided by the driver.

The device MUST NOT reuse a command ID until it has received an output
buffer containing VIRTIO_BALLOON_CMD_ID_STOP from the driver.

The device MUST ignore pages that are provided with a command ID that does
not match the current value in \field{free_page_hint_cmd_id}.

If the content of a previously hinted page has not been modified by the
guest since the device issued the \field{free_page_hint_cmd_id} associated
with the hint, the device MAY modify the contents of the page.

The device MUST NOT modify the content of a previously hinted page after
\field{free_page_hint_cmd_id} is set to VIRTIO_BALLOON_CMD_ID_DONE.

The device MUST report a value of VIRTIO_BALLOON_CMD_ID_DONE in
\field{free_page_hint_cmd_id} when it no longer has need for the
previously hinted pages.

\paragraph{Legacy Interface: Free Page Hinting}\label{sec:Device Types / Memory Balloon Device / Device Operation / Free Page Hinting / Legacy Interface: Free Page Hinting}

When using the legacy interface, transitional devices and drivers MUST
format the command ID field in output buffers according to the native
endian of the guest rather than (necessarily when not using the legacy
interface) little-endian.

\subsubsection{Page Poison}\label{sec:Device Types / Memory Balloon Device / Device Operation / Page Poison}

Page Poison provides a way to notify the host that the guest is initializing
free pages with \field{poison_val}. When the feature is enabled, pages will
be immediately written to by the driver after deflating, and pages reported
by free page reporting will retain the value indicated by \field{poison_val}.

If the guest is not initializing freed pages, the driver should reject the
VIRTIO_BALLOON_F_PAGE_POISON feature.

If VIRTIO_BALLOON_F_PAGE_POISON feature has been negotiated, the driver
will place the initialization value into the \field{poison_val}
configuration field data.

\drivernormative{\paragraph}{Page Poison}{Device Types / Memory Balloon Device / Device Operation / Page Poison}

Normative statements in this section apply if the
VIRTIO_BALLOON_F_PAGE_POISON feature has been negotiated.

The driver MUST initialize the deflated pages with \field{poison_val} when
they are reused by the driver.

The driver MUST populate the \field{poison_val} configuration data before
setting the DRIVER_OK bit.

The driver MUST NOT modify \field{poison_val} while the DRIVER_OK bit is set.

\devicenormative{\paragraph}{Page Poison}{Device Types / Memory Balloon Device / Device Operation / Page Poison}

Normative statements in this section apply if the
VIRTIO_BALLOON_F_PAGE_POISON feature has been negotiated.

The device MAY use the content of \field{poison_val} as a hint to guest
behavior.

\subsubsection{Free Page Reporting}\label{sec:Device Types / Memory Balloon Device / Device Operation / Free Page Reporting}

Free Page Reporting provides a mechanism similar to balloon inflation,
however it does not provide a deflation queue. Reported free pages can
be reused by the driver after the reporting request has been acknowledged
without notifying the device.

The driver will begin reporting free pages. When exactly and which free
pages are reported is up to the driver.

\begin{enumerate}

\item The driver determines it has enough pages available to begin
  reporting free pages.

\item The driver gathers free pages into a scatter-gather list and adds
  them to the reporting_vq.

\item The device acknowledges the reporting request by using the
  reporting_vq descriptor.

\item Once the device has acknowledged the report, the driver can reuse the
  reported free pages when needed (e.g., by putting them back to free page
  lists in the guest operating system).

\item The driver can then continue to gather and report free pages until it
  has determined it has reported a sufficient quantity of pages.

\end{enumerate}

\drivernormative{\paragraph}{Free Page Reporting}{Device Types / Memory Balloon Device / Device Operation / Free Page Reporting}

Normative statements in this section apply if the
VIRTIO_BALLOON_F_PAGE_REPORTING feature has been negotiated.

If the VIRTIO_BALLOON_F_PAGE_POISON feature has not been negotiated, then
the driver MUST treat all reported pages as uninitialized memory.

If the VIRTIO_BALLOON_F_PAGE_POISON feature has been negotiated, the
driver MUST initialize all free pages with \field{poison_val} before
reporting them.

The driver MUST NOT use the reported pages until the device has
acknowledged the reporting request.

The driver MAY report free pages any time after DRIVER_OK is set.

The driver SHOULD attempt to report large pages rather than smaller ones.

The driver SHOULD avoid reading/writing reported pages if
not strictly necessary.

\devicenormative{\paragraph}{Free Page Reporting}{Device Types / Memory Balloon Device / Device Operation / Free Page Reporting}

Normative statements in this section apply if the
VIRTIO_BALLOON_F_PAGE_REPORTING feature has been negotiated.

If the VIRTIO_BALLOON_F_PAGE_POISON feature has not been negotiated, the
device MAY modify the contents of any page supplied in a report request
before acknowledging that request by using the reporting_vq descriptor.

If the VIRTIO_BALLOON_F_PAGE_POISON feature has been negotiated, the device
MUST NOT modify the the content of a reported page to a value other than
\field{poison_val}.
